Modeling the Evolution of Mimicry

### 2016-02-26 docs/paper/Paper.tex
12 pages...!!!

In developing Echo using Holland's criterion, a step by step approach has been taken. Among these steps of improvement, the work by Hraber \citep{hraber1997} is more from an ecological perspective, and having a good resemblance with natural ecosystem. It has stable patterns of interactions among genetic variants, or trophic networks. It consists of different levels of resources along with relative species abundance. Even different cataclysmic events such as meteor, drought and so fourth are introduced. Also several isolation effects are observed where a number of sites are allowed to evolve independently. In addition to that Echo contains interesting flow of resources.

Two separate experiments were done to observe whether species abundance pattern can be used to find ecologically plausible behavior in Echo. To make a falsifiable comparison with biological systems, two ecological patterns are considered, each of which has similar quantitative properties in many different ecosystems that exist in nature. The objective is to compare these naturally occurring ecosystems with Echo. The ecological patterns are the Preston  curve and the species-area scaling relation. This allowed comparison of Echo diversity patterns with empirical patterns of species abundance. As result of the comparison Echo showed good quantitative agreement with naturally occurring species abundance distribution and species area scaling relation.

@@

Repulsive animals, such as heliconiids are very conspicuously colored. Having this noticeable property, they are easily recalled by predators. Their wing pattern works as a warning to predators. Once a predator has the knowledge of their inedible and unpalatable property, they would probably never attempt to try it again. As this is true, if any organism within close family and species, but being edible and having a deceptive resemblance to those conspicuously colored species will be avoided by the predators. Wickler \citep{wickler1986} expresses,
\begin{quote}
\textsl{``Such unpalatable appearing and yet edible animals thus possess a false warning pattern, they `act a part'. An actor is a mime, and so the representation of a false warning pattern was called \textit{mimicry}. Since Bates was the first to point out this phenomenon, it has received the name \textit{Batesian Mimicry} in his honor."}
\end{quote}
In general, the animal which is avoided by predator for unpalatable behavior is called the \textbf{model} and the imitating animal is called the \textbf{mimic}.

@@

Mimicry is one of the most fascinating cases of evolutionary biology. This phenomenon has been studied extensively as infinite number of evidence exists in nature. One hundred years after Bates first clearly defined the concept, a review of literature listed 1,500 papers arguing for or against it. And this figure is more than 40 years old as it has been taken from Wilcker \citep{wickler1986}, published in 1968. It is almost impossible to survey all the known examples of Mimicry, and this would in any case be pointless. Then again many examples can be found in Cott's \textit{Adaptive Coloration in Animals} \citep{cott1957}.

@@

\subsection{FormAL Framework}
The \textsl{``FormAL framework"} is a collection of ideas and concepts taken from Peter Grogono's \gls{formal} project \citep{grogono2003} and are used to build a framework for modeling the evolution of mimicry. The primary goal of the FormAL project was ``\textit{To study the emergence of complexity}". While the principal behind it was ``\textit{not to include a variable in an agent unless the variable is genetically controlled (or, at least, genetically influenced)}". We will see that this principal was not completely established in simulating mimicry as it was not possible to come up with a genetic representation of Hopfield Network which was used as a pattern recognizer for the predator species.

@@

\paragraph{Cell}
The entire three dimensional toroidal environment is divided in multiple cells. A cell is a three dimensional cubical section of the hyperspace. Accordingly in table \ref{tab:environment-control-parameters} \(ISize\) is the parameter that controls the number of cells in the environment. The size of a cell is evaluated from the entire world size. And the total number of cell is calculated as \(ISize^3 = 216\).

@@

\subsubsection{Learning}
The objective of a predator's interaction with prey is always to consume it. But based on the prey's pattern and palatability, the predator will either be able to consume it or throw it back to the environment. At this event the predator needs to learn the pattern with which the prey has been represented. The pattern represents palatability of the prey species, at least to the predator. To store this new pattern into memory, each predator is equipped with a Hopfield Network Memory. Every time a new interaction is made by the predator its memory is initialized with all the existing pattern that has already been encountered and the new one. The learning procedure used for this memory is Hebbian Learning \citep{hebb1949}.

The following \textit{general learning rule} is adopted in neural network studies: \textit{The weight vector} \( \textbf{w}_i = [w_{i1} \> w_{i2} \> ... \> w_{in}]^t \) \textit{increases in proportion to the product of input} \textbf{x} \textit{and learning signal r}. The learning signal r is in general a function of \(\textbf{w}_i,\textbf{x}\), and sometimes of the teacher's signal \(d_i\). Thus we have, 
\begin{equation}
	r = r(\textbf{w}_i,\textbf{x},d_i)
%\label{eq:}
\end{equation}
The increment of the weight vector \(\textbf{w}_i\) produced by the learning step at time t according to the general learning rule is,
\begin{equation}
	\Delta\textbf{w}_i(t)=cr[\textbf{w}_i(t),\textbf{x}(t),d_i(t)]\textbf{x}(t)
%\label{eq:}
\end{equation}
when c is a positive number called the \textit{learning constant} that determines the rate of learning.

For Hebbian learning rule the learning signal is equal simply to the neuron's output. We have
\begin{equation}
	r \overset{\Delta}{=} f(\textbf{w}_i^t \textbf{x})
%\label{eq:}
\end{equation}
The increment \(\Delta\textbf{w}_i\) of the weight vector becomes
\begin{equation}
	\Delta\textbf{w}_i = cf(\textbf{w}_i^t \textbf{x})\textbf{x}
%\label{eq:}
\end{equation}
The single weight \( w_{ij} \) is adapted using the following increment:
\begin{equation}
	\Delta\textit{w}_{ij} = cf(\textbf{w}_i^t \textbf{x})x_j
%\label{eq:}
\end{equation}
This can be written briefly as 
\begin{equation}
	\Delta\textit{w}_{ij} = c o_i x_j, \> \text{for} \> j = 1, 2, ..., n
%\label{eq:}
\end{equation}
Here, \(\textbf{o}\) is the output vector. The learning rule requires the weight initialization at small random values around \( \textbf{w}_i = \textbf{0}\) prior to learning. The Hebbian learning rule represents a purely feed-forward, unsupervised learning.

Training for Hopfield Network used for predator's memory is done by Hebbian Learning. Initially the weights are all set to zero. Using Hebbian rule, the outer product of the input - output vector pairs are calculated for each pattern. As Hopfield is a feedback network, the output of the network is also the input. The outer vector matrix of all the patterns are summed to come up with the final weight matrix. Each component of the weight matrix \(\textbf{W} = \{w_{ij}\}\) is given by:
\begin{equation}
w_{ij} = \sum_{p=1}^{P} s_i(p) t_j(p), i \neq j
%\label{eq:}
\end{equation}
\[
w_{ij} = 0, i = j
\]
where P is the number of patterns. Vectors \textbf{S} and \textbf{T} are respectively, the input and the desired output of the network.

\subsubsection{Design of Memory with Hopfield Network}

\paragraph{Hopfield Network}

Algorithm for the Hopfield model is described in the following:

\begin{enumerate}
	\item \textit{Learning.} Let \( \boldsymbol{\xi}_1, \boldsymbol{\xi}_2, ..., \boldsymbol{\xi}_{\mu} \) denote a known set of N-dimensional fundamental memories. Use the outer-product rule (i.e. Hebb's postulate of learning) to compute the synaptic weights of the network as 
	\begin{equation}
		w_{ji} =
			\begin{cases}
				\frac{1}{N} \sum_{\mu=1}^{M}\xi_{\mu,j}\xi_{\mu,i}	& j \neq i \\
				0																										& j = i 
			\end{cases}
	%\label{eq:}
	\end{equation}
	where \(w_{ji}\) is the synaptic weight from neuron \(i\) to neuron \(j\). The elements of the vector \( \boldsymbol{\xi}_\mu \) equal \(\pm 1\) (bipolar). Once they are computed, the synaptic weights are kept fixed.

	\item \textit{Initialization.} Let \( \boldsymbol{\xi}_{probe} \) denote an unknown N-dimensional input vector (probe) presented to the network. The algorithm is initialized by setting
	\begin{equation}
		x_j(0) = \xi_{j,probe}, \> j = 1,...,N
	%\label{eq:}
	\end{equation}
	where \( x_j(0) \) is the state of neuron \(j\) at time \(n = 0\) and \( \xi_{j,probe} \) is the \(j\)th element of the probe \( \boldsymbol{\xi}_{probe} \).

	\item \textit{Iterate Until Convergence.} Update the elements of state vector \( \boldsymbol{x}(n) \) asynchronously (i.e., randomly and once at a time) according to the rule
	\begin{equation}
		x_j(n+1)=sgn \Bigg(\sum_{i=1}^{N} w_{ji}x_i(n) \Bigg), \> j = 1,2, ..., N
	%\label{eq:}
	\end{equation}
	Repeat the iteration until the state vector \( \boldsymbol{x} \) remains unchanged.

	\item \textit{Output.} Let \( \boldsymbol{x}_{fixed} \) denote the fixed point (stable state) computed at the end of the step 3, The resulting output vector \( \boldsymbol{y} \) of the network is
	\begin{equation}
		\boldsymbol{y} = \boldsymbol{x}_{fixed}
	\label{eq:}
	\end{equation}
	Step 1 is the storage phase, and steps 2 through 4 constitute the retrieval phase.
\end{enumerate}

@@

The predators are distributed all over the cells with a constant mobile behavior to switch to another neighboring cell depending on the one containing the least number of predator and also most number of prey species. This behavior of predator has been designed to enforce the predatory behavior of these agents and also to have increased predator prey interaction in the simulation in terms of one agent chasing the other for survival of species.

@@

\begin{table}[h!]
\centering
\setlength\tabcolsep{2pt}
\begin{tabular}{| p{2cm} | >{\centering} p{1cm} | p{5cm} |}
	\hline
		\textbf{Parameter} & \textbf{Value} & \textbf{Description} \\ \hline
		Minimum Memory Size & 2 to 6 & Minimum number of patterns stored in memory before predators start making intelligent decisions.\\ \hline
		Maximum Memory Size & 10 & Maximum number of patterns to be stored in memory. Limited to reduce processing time. \\ \hline 
		Hopfield Maximum Iterations & 20 & Maximum number of iterations for Hopfield Network to recognize a pattern. Usually the network reaches a steady state before that. But this restriction is to avoid infinite loop in case the network never reaches a steady state. \\ \hline
		Attack Age & 500 & minimum age a predator needs to reach to be able to attack prey species.  \\ \hline
		Attack Interval & 100 & Interval of time which needs to pass before a predators attacks its next prey. \\ \hline
		Genome Mutation rate & 0.3 & Mutation rate for the 5 bit genome of the predators representing their mobility and pattern recognition capability. \\ \hline
		Reproduction Age Limit & 500 & Minimum age a predator needs to reach before engaging in reproduction.\\ \hline
		Reproduction Interval & 1000 to 3000 & Interval of time a predator needs to pass between two reproduction process.\\ \hline
		Demise Age & 2000 to 7000 & Age at which a predator is considered as dead.\\
	\hline
\end{tabular}
\caption{Parameters to control predator population and pattern recognition capability.}
\label{tab:predator-control-parameters}
\end{table}

@@

If we consider the field of Artificial Life as a tool for biological inquiry then the model presented in this thesis was successful in terms of providing good insight in terms of understanding better the evolutionary process of mimicry as it happens in nature. As mentioned previously this model gives us verification of the theory of Turner in explaining the evolution of mimicry with punctuated equilibrium. Also it enforces findings from the work of Franks and Noble that multiple Mullerian mimics do not converge into one large ring.

If we consider Artificial Life as the study where we simulate natural processes to further extend capabilities in the field of computer science, then also this thesis is a success in making an appropriate emulation of the evolutionary process of mimicry. We can conclude that the complex behavior of Batesian and Mullerian mimicry can be simulated with this model.

>>>>

In developing Echo using Holland's criterion, a step by step approach has been taken. Among these steps of improvement, the work by Hraber \citep{hraber1997} is more from an ecological perspective, having a good resemblance with natural ecosystem.

@@

Repulsive animals, such as heliconiids are very conspicuously colored. Having this noticeable property, they are easily recalled by predators. Their wing pattern works as a warning to predators. Once a predator has the knowledge of their inedible and unpalatable property, they would probably never attempt to try it again. As this is true, if any organism within close family and species, but being edible and having a deceptive resemblance to those conspicuously colored species will be avoided by the predators.

In general, the animal which is avoided by predator for unpalatable behavior is called the \textbf{model} and the imitating animal is called the \textbf{mimic}.

@@

@@

\subsection{FormAL Framework}
The \textsl{``FormAL framework"} is a collection of ideas and concepts taken from Peter Grogono's \gls{formal} project \citep{grogono2003} and are used to build a framework for modeling the evolution of mimicry. The primary goal of the FormAL project was ``\textit{To study the emergence of complexity}". While the principal behind it was ``\textit{not to include a variable in an agent unless the variable is genetically controlled (or, at least, genetically influenced)}".

@@

\paragraph{Cell}
The entire three dimensional toroidal environment is divided in multiple cells. A cell is a three dimensional cubical section of the hyperspace. Accordingly in Table \ref{tab:environment-control-parameters} \(ISize\) is the parameter that controls the number of cells in the environment. The size of a cell is evaluated from the entire world size. And the total number of cell is calculated as \(ISize^3 = 216\).

@@

\subsubsection{Learning}
The objective of a predator's interaction with prey is always to consume it. But based on the prey's pattern and palatability, the predator will either be able to consume it or throw it back to the environment. At this event the predator needs to learn the pattern with which the prey has been represented. The pattern represents palatability of the prey species, at least to the predator. Every time a new interaction is made by the predator its memory is initialized with all the existing pattern that has already been encountered and the new one. The learning procedure used for this memory is Hebbian Learning \citep{hebb1949}, which represents a purely feed-forward, unsupervised learning. Initially the weights of the Hopfield Network are all set to zero. Using Hebbian rule, the outer product of the input - output vector pairs are calculated for each pattern. As Hopfield is a feedback network, the output of the network is also the input. The outer vector matrix of all the patterns are summed to come up with the final weight matrix.

@@

The predators are distributed all over the cells with a constant mobile behavior to switch to another neighboring cell depending on the one containing the least number of predator and also most number of prey species. This behavior of predator has been designed to enforce the predatory capabilities of these agents and also to have increased predator prey interaction in the simulation in terms of one agent chasing the other for survival of species.

@@

\begin{table}[h!]
\centering
\setlength\tabcolsep{2pt}
\begin{tabular}{| p{2cm} | >{\centering} p{1cm} | p{5cm} |}
	\hline
		\textbf{Parameter} & \textbf{Value} & \textbf{Description} \\ \hline
		Minimum Memory Size & 2 to 6 & Minimum number of patterns stored in memory before predators start making intelligent decisions.\\ \hline
		Maximum Memory Size & 10 & Maximum number of patterns to be stored in memory. Limited to reduce processing time. \\ \hline 
		Hopfield Maximum Iterations & 20 & Maximum number of iterations to recognize a pattern and to avoid infinite loop in case the network never reaches a steady state. \\ \hline
		Attack Age & 500 & minimum age a predator needs to reach to be able to attack prey species.  \\ \hline
		Attack Interval & 100 & Interval of time which needs to pass before a predators attacks its next prey. \\ \hline
		Genome Mutation rate & 0.3 & Mutation rate for the 5 bit genome of the predators representing their mobility and pattern recognition capability. \\ \hline
		Reproduction Age Limit & 500 & Minimum age a predator needs to reach before engaging in reproduction.\\ \hline
		Reproduction Interval & 1000 to 3000 & Interval of time a predator needs to pass between two reproduction process.\\ \hline
		Demise Age & 2000 to 7000 & Age at which a predator is considered as dead.\\
	\hline
\end{tabular}
\caption{Parameters to control predator population and pattern recognition capability.}
\label{tab:predator-control-parameters}
\end{table}

@@
